CRAN packages network graph
========================================================

Part of a project for Network analysis class is constructing package dependency tree off of CRAN (I called shotgun). I could scrape CRAN website for packages and their dependencies, suggests, imports and enhancements ([as done here](http://librestats.com/2012/05/17/visualizing-the-cran-graphing-package-dependencies/)), but I opted for a more elegant solution via `available.packages` function, that reports which packages are available on CRAN and also adds all the above relationships. Feel free to add to this function by pulling [my GitHub repo](https://github.com/romunov/rdepnet).

```{r}
library(igraph)
library(reshape2)
library(stringr)
library(RColorBrewer)
```

```{r}
# Get a list of available packages.
dep.data <- available.packages(contriburl = "http://cran.at.r-project.org/bin/windows/contrib/3.0")
dep.data[1:5, 1:5]
# Get a subset of variables of interest.
dep.data <- dep.data[, c("Package", "Depends", "Imports", "Suggests", "Enhances")]

# Some (most?) packages depends/import to multiple packages. Let's 
# tease those packages apart.
# This function will make a "long" list of relationships
# for each package (where applicable). If there's no relationship, the output is "none".
make.data <- apply(dep.data, 1, FUN = function(x) {

      if (all(is.na(x[-1]))) return(data.frame(from = x[1], to = x[1], relation = "none"))
      
      x2 <- na.omit(x[-1])
      x2 <- sapply(x2, strsplit, ",")
      
      out <- data.frame(from = rep(x[1], length(unlist(x2))), to = unlist(x2), relation = rep(names(x2), times = sapply(x2, length)))
      rownames(out) <- 1:nrow(out)
      out
   })
done.data <- do.call("rbind", make.data)
rownames(done.data) <- 1:nrow(done.data)

# We clean the data. 
clean.data <- done.data # make a copy
# remove version numbers
clean.data$to <- sub("\\((.*?)\\)", "", x = clean.data$to, perl = TRUE)
# remove newlines
clean.data$to <- sub("\\\n", "", x = clean.data$to, perl = TRUE)
# remove dependency on R (*)
clean.data[clean.data$to == "R" | clean.data$to == "R ", "to"] <- as.character(clean.data[clean.data$to == "R" | clean.data$to == "R ", "from"])
# clean whitespace
clean.data$to <- as.factor(str_trim(clean.data$to, "both"))
# remove loops
clean.data <- clean.data[!apply(clean.data[, c("from", "to")], 1, function(x) x[1] == x[2]), ] 
```

This is the function that will be doing the heavy drawing for us.
```{r}
#' Draws a subset of a data.frame as a network graph.
#' @param x data.frame. Has at least three columns, from, to and relation.
#' @param relation Character vector of length one. Rows with what relation from column relation (see x) should be used?
#' @param pkg Character. Package name to be subsetted.
#' @param ... Arguments passed to plot function that plots igraph object.
#' @return \code{igraph} object with the possible side effect of plotting the object.

plotGraph <- function(x, relation = NULL, pkg.to = NULL, 
   pkg.from = NULL, plot = TRUE, ...) {
   
   # prepare the data to be plotted
   if (!is.null(relation)) {
      gd <- droplevels(x[x$relation %in% relation, ])
   } else {
      gd <- x
   }
   if (!is.null(pkg.to)) gd <- gd[gd$to %in% pkg.to, ]
   if (!is.null(pkg.from)) gd <- gd[gd$from %in% pkg.from, ]
   
   # make the data an igraph object and plot it (if specified)
   g <- graph.data.frame(gd, directed = TRUE)
   
   cc <- E(g)$relation
   if (any(unique(cc) == "Imports")) cc[cc == "Imports"] <- "#E41A1C"
   if (any(unique(cc) == "Suggests")) cc[which(cc == "Suggests")] <- "#984EA3"
   if (any(unique(cc) == "Enhances")) cc[which(cc == "Enhances")] <- "#4DAF4A"
   if (any(unique(cc) == "Depends")) cc[which(cc == "Depends")] <- "#377EB8"
   E(g)$color <- cc
   
   if (plot) plot(g, ...)
   # see ?igraph:::layout for possible layout options
   g
}
```

We can try and draw a full dependency graph, but it doesn't look pretty. It also takes a long time. Consider before executing this next chunk. Output of `plotGraph` function is an `igraph` object, which is why we save it to a variable for possible further manipulation.
```{r, eval = FALSE}
graph.all.depends <- plotGraph(x = clean.data, relation = "Depends",
   vertex.label.color = "blue", edge.width = 0.5, edge.color = "#ACC49D",
   edge.arrow.size = 0.5, vertex.label.cex = 0.7, vertex.size = 0, 
   vertex.color = "white", layout = layout.kamada.kawai)
```

Another example of usage would be to plot the relationships between packages for high performance computing (see [HPC task view](http://cran.r-project.org/web/views/HighPerformanceComputing.html) on CRAN).

```{r}
hpp <- c("batch", "BatchExperiments", "BatchJobs", "bayesm", "bcp", 
"biglars", "biglm", "bigmemory", "bigrf", "biopara", "bnlearn", 
"caret", "cudaBayesreg", "data.table", "dclone", "doMC", "doMPI", 
"doRedis", "doRNG", "doSNOW", "ff", "foreach", "fork", "GAMBoost", 
"gcbd", "Geneland", "gputools", "GridR", "HadoopStreaming", "harvestr", 
"HiPLARM", "inline", "latentnet", "lga", "magma", "mapReduce", 
"Matching", "mchof", "multicore", "nws", "OpenCL", "orloca", 
"partDSA", "pbdBASE", "pbdDEMO", "pbdDMAT", "pbdMPI", "pbdNCDF4", 
"pbdSLAP", "peperr", "permGPU", "pmclust", "profr", "proftools", 
"pvclust", "Rcpp", "Rdsm", "rgenoud", "RInside", "rJava", "rlecuyer", 
"Rmpi", "rpud", "rredis", "rsprng", "scaleboot", "snow", "snowfall", 
"snowFT", "speedglm", "sprint", "sqldf", "STAR", "tm", "varSelRF", 
"WideLM", "xgrid")

graph.parallel.depends <- plotGraph(x = clean.data, pkg.to = hpp, relation = "Depends",
   vertex.label.color = "black", edge.width = 1, edge.arrow.size = 0.2, 
   vertex.label.cex = 0.9, vertex.size = 0, vertex.color = "white", 
   layout = layout.kamada.kawai)
```

Or perhaps we'd be interested in seeing relationships for parallel computing packages. We notice that `foreach`, `snow` and `Rmpi` are quite popular. `multicore` and `snowfall` (this one being my fav) have less relationships.

Oh, I forgot to mention, lines are color coded and denote different relationship types (colors were taken from `brewer.pal(4, "Set1")`).
```{r}
graph.somepara.all <- plotGraph(x = clean.data, 
      pkg.to = c("snowfall", "snow", "foreach", "Rmpi", "multicore", 
      "sprint", "biopara", "bigrf", "doMPI", "doRedis", "nws", "snowFT"), 
   vertex.label.color = "black", edge.width = 1, edge.arrow.size = 0.2, 
   vertex.label.cex = 0.9, vertex.size = 0, vertex.color = "white", 
   layout = layout.kamada.kawai)
```

Let's see only depends for parallel computing packages. Notice that depends are colored blue(ish). Quite a few packages depend on `foreach`.
```{r}
graph.somepara.depends <- plotGraph(x = clean.data, 
      pkg.to = c("snowfall", "snow", "foreach", "Rmpi", "multicore", 
      "sprint", "biopara", "bigrf", "doMPI", "doRedis", "nws", "snowFT"), 
   relation = "Depends",
   vertex.label.color = "black", edge.width = 1, edge.arrow.size = 0.2, 
   vertex.label.cex = 0.9, vertex.size = 0, vertex.color = "white", 
   layout = layout.kamada.kawai)
```

I also fancy packages for spatial data. Barry has made a contribution on this subject, and you can [find the graphs here](http://www.maths.lancs.ac.uk/~rowlings/Teaching/UseR2012/introduction2Talk.html).
A lot of packages depend on `sp`.

```{r}
graph.sp2 <- plotGraph(x = clean.data, pkg.to = "sp", relation = "Depends",
   vertex.label.color = "black", edge.width = 0.1, edge.color = "#377EB8",
   vertex.label.cex = 0.9, vertex.size = 0, 
   vertex.color = "white", layout = layout.kamada.kawai)
```

```{r}
spp <- c("ade4", "adehabitat", "adehabitatHR", "adehabitatHS", "adehabitatLT", 
"adehabitatMA", "ads", "akima", "ash", "aspace", "automap", "classInt", 
"CompRandFld", "constrainedKriging", "cshapes", "DCluster", "deldir", 
"DSpat", "ecespa", "fields", "FieldSim", "gdistance", "Geneland", 
"GEOmap", "geomapdata", "geonames", "geoR", "geoRglm", "geosphere", 
"geospt", "geospt", "GeoXp", "ggmap", "glmmBUGS", "gmt", "GriegSmith", 
"gstat", "Guerry", "hdeco", "intamap", "landsat", "mapdata", 
"mapproj", "maps", "maptools", "MarkedPointProcess", "MBA", "Metadata", 
"micromap", "ModelMap", "ncdf", "ncf", "nlme", "OpenStreetMap", 
"osmar", "pastecs", "PBSmapping", "PBSmodelling", "plotKML", 
"psgp", "ramps", "RandomFields", "rangeMapper", "RArcInfo", "raster", 
"rasterVis", "RColorBrewer", "regress", "rgdal", "rgeos", "RgoogleMaps", 
"RPyGeo", "RSAGA", "RSurvey", "rworldmap", "sgeostat", "shapefiles", 
"sp", "spacetime", "sparr", "spatcounts", "spatgraphs", "spatial", 
"spatialCovariance", "SpatialExtremes", "spatialkernel", "spatialprobit", 
"spatialsegregation", "spatstat", "spBayes", "spcosa", "spdep", 
"spgrass6", "spgwr", "sphet", "splancs", "spsurvey", "Stem", 
"tgp", "trip", "tripack", "tripEstimation", "UScensus2000blkgrp", 
"UScensus2000cdp", "UScensus2000tract", "vardiag", "vec2dtransf", 
"vegan")
graph.spp <- plotGraph(x = clean.data, pkg.to = spp, relation = "Depends",
   vertex.label.color = "black", edge.width = 0.1, edge.color = "#377EB8",
   vertex.label.cex = 0.5, vertex.size = 0, edge.arrow.size = 0.2,
   vertex.color = "white", layout = layout.kamada.kawai)
```

What a mess! Let's see what's going on for the top few packages.

```{r}
top.sp <- names(sort(degree(graph.spp), decreasing = TRUE)[1:20])
graph.top.spp <- plotGraph(x = clean.data, pkg.to = top.sp, relation = "Depends",
   vertex.label.color = "black", edge.width = 0.1, edge.color = "#377EB8",
   vertex.label.cex = 0.5, vertex.size = 0, edge.arrow.size = 0.2,
   vertex.color = "white", layout = layout.kamada.kawai)
```

Better, but let's narrow it down even further. We exclude `nlme` and `RColorBrewer`.

```{r}
top.sp <- names(sort(degree(graph.spp), decreasing = TRUE)[1:12])
top.sp <- top.sp[which(!top.sp %in% c("nlme", "RColorBrewer"))]
graph.top.spp <- plotGraph(x = clean.data, pkg.to = top.sp, relation = "Depends",
   vertex.label.color = "black", edge.width = 0.1, edge.color = "#377EB8",
   vertex.label.cex = 0.6, vertex.size = 0, edge.arrow.size = 0.2,
   vertex.color = "white", layout = layout.kamada.kawai)
```

It's up to you to explore your favorite packages' networks further. For comments or questions, feel free to contact me.